Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-16 appendix/appendix.tex
New content
Grammar fix

\begin{table}[H]
    \centering
    \begin{tabular}{|c c|}
        \hline
        Environment                 & Version                       \\ [0.5ex]
        \hline\hline
        Placeholder                 & 2\\
        \hline
    \end{tabular}
    \caption{Environment on remote machine}
    \label{table:remote_macine}
\end{table}

\section{Data analysis}

\chapter{Data and settings}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c|}
        \hline
        Parameter            & Value \\[0.5ex]
        \hline\hline
        $\lambda_N$          & 2.20  \\
        $\kappa_{N\beta}$    & 3.96  \\
        $\mu_N$              & 1.72  \\
        $\nu_{N\Phi}$        & 0.219 \\
        \hline
        $\lambda_\Phi$       & 1.31 \\
        $\kappa_{\Phi\beta}$ & 0.124 \\
        $\mu_\Phi$           & 0.145 \\
        \hline
        $s_{\beta N}$        & 6.55  \\
        $i_{\beta\Phi}$ & 1.71 \\
        $\mu_\beta$          & 0.521  \\
        \hline
        $s_{\alpha\Phi}$     & 10.2 \\
        $\mu_\alpha$         & 19.7  \\
        \hline
    \end{tabular}
    \caption[TODO]
    {Known values TODO}
    \label{table:known_values}
\end{table}

@@

    \caption[Acceptance rates of different kernels, using median epsilon strategy]
    {Acceptance rates of different kernels, using median epsilon strategy. Each population has 2000 particles}
    \label{fig:acceptance2}

@@

    \caption{Total sampling size}
    \label{fig:resultCurve_uni}

@@

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/inferno5.png}}
    \end{center}
    \caption[Joint density distribution of each parameter pair]%
    {Joint distribution of each parameter pair. MORE DESCRIPTION}
    \label{fig:model5_para_joint}
\end{figure}

>>>>

Cirrus hardware can be found in \url{https://www.cirrus.ac.uk/about/hardware.html}. The software environment used in the development and experiments are listed in Table \ref{table:cirrus_soft}. Requirement of site packages is the sanme as that on local macine (Table \ref{table:local_package})

\begin{table}[H]
    \centering
    \begin{tabular}{|c c|}
        \hline
        Environment           & Version                      \\ [0.5ex]
        \hline\hline
        Operating system      & Red Hat Enterprise Linux 8.1 \\
        miniconda environment & 4.8.3                        \\
        Python interpreter    & 3.7.7                        \\
        gcc                   & 6.3.0                        \\
        \hline
    \end{tabular}
    \caption{Environment on local machine}
    \label{table:cirrus_soft}
\end{table}

% \section{Data analysis}

\chapter{Data and supplementary figures}

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c|}
        \hline
        Parameter            & Value \\[0.5ex]
        \hline\hline
        $\lambda_N$          & 2.20  \\
        $\kappa_{N\beta}$    & 3.96  \\
        $\mu_N$              & 1.72  \\
        $\nu_{N\Phi}$        & 0.219 \\
        \hline
        $\lambda_\Phi$       & 1.31  \\
        $\kappa_{\Phi\beta}$ & 0.124 \\
        $\mu_\Phi$           & 0.145 \\
        \hline
        $s_{\beta N}$        & 6.55  \\
        $i_{\beta\Phi}$      & 1.71  \\
        $\mu_\beta$          & 0.521 \\
        \hline
        $s_{\alpha\Phi}$     & 10.2  \\
        $\mu_\alpha$         & 19.7  \\
        \hline
    \end{tabular}
    \caption[Parameter values used to generate synthetic data for model 1]
    {Parameter values used to generate synthetic data for model 1. They were obtained from a preliminarily least-square fitting}
    \label{table:known_values}
\end{table}

@@

    \caption[Acceptance rates of different kernels, using median epsilon strategy]
    {Acceptance rates of different kernels, using median epsilon strategy. Each population has 2000 particles. The total length of populations is different because some runs reached the final $\epsilon_T$ earlier than others}
    \label{fig:acceptance2}

@@

    \caption{Simulated data from the last population of model 1, 2 and 3, using uniform prior.  Error bar indicates SEM (for $N$ and $\Phi$ the SEM is relatively small)}
    \label{fig:resultCurve_uni}

@@

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/inferno5.png}}
    \end{center}
    \caption[Joint density distribution of each parameter pair]%
    {Joint distribution of each parameter pair. The diagonal is KDE fit of each individual parameter, the upper triangle contains scatter plot of the particles and the lower triangle is the joint density distribution for each parameter pair}
    \label{fig:model5_para_joint}
\end{figure}

\begin{figure}[ht]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/credible.png}}
    \end{center}
    \caption[Credible intervals of parameters in model 5]%
    {Credible intervals of parameters in model 5 across all generations}
    \label{fig:credible}
\end{figure}

### 2020-08-16 background/background.tex
Grammar fix

Compared to mammals, zebrafish can regenerate spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches due to this ability. Experiments found that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al.\cite{ref:Tsarouchas} studied the the complex interactions happen at the lesion site. The inflammation affects the regeneration, as if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration\cite{ref:Tsarouchas}. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, are related the inflammation, and they are dynamically controlled by immunes cells, i.e. immune cells produce and control the cytokines-mediated inflammation \cite{ref:Tsarouchas}. Cytokines are small proteins, work as messenger to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map (Figure \ref{fig:map}) can be drawn according to the evidences found in experiments.\cite{ref:Tsarouchas}. After injury, a massive immune response was observed, with the increasing numbers of several kinds of cells. This dynamical system consists of complex interactions and dynamical feedback between cells and cytokines, For example, il-1$\beta$ promotes macrophages' increase, and as a major source, more macrophages leads to more expression of tnf-$\alpha$ mRNA.

Consequently, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration. il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect; tnf-$\alpha$ promotes the regeneration all the time. The presence of the cytokines also affects the number of immune cells, while as sources, more immune cells produce more cytokines. Interactions between different types of cells are also considered in the system (e.g., the number of neutrophil is affected by macrophages).

@@

    \caption[Working model of the influence of the innate immune system on axonal regrowth]%
    {Working model of the influence of the innate immune system on axonal regrowth, from Tsarouchas et al.\cite{ref:Tsarouchas}. Lines ended with arrow represent promoting effect, lines ended with T-connectors represent inhibition}
    \label{fig:map}

@@

This dynamic system is suitable for mathematical modelling, where a set of equations can be written to represent the dynamics of different objects e.g. each types of cells and cytokines. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions inside the system, based on the proposed hypothesis \cite{ref:Tsarouchas}, and furthermore based our own hypothesis which may include different interactions.

The biological background of works in this dissertation is mostly related the the experimental evidences, quantitative data and proposed hypothesis in \cite{ref:Tsarouchas}. The published data on the quantitative measurement of cells and cytokines make the modelling and further analysis possible. Some other interactions proposed in our modelling stages are enlightened from Anderson et al. \cite{Anderson}, where intracellular cytokine signaling networks were established and computationally modelled.

@@

Mathematical models can express real-life problems in mathematical forms, and provide qualitative or quantitative inside-look of the question through various mathematical tools and techniques. Mathematical models can be used for prediction, classification, testing our understanding or hypothesis and many other tasks, thus they are widely used to solve practical problems. However, for a model that describes the zebrafish spinal cord regeneration process, we were less interested to the model's predictive functionality; the main focus is on the explore of uncertainty in parameter estimates and comparison with alternative models. In this case, we wish to use modelling techniques to implement inference methods, test our hypothesis and hopefully find and compare well-fitted models.

According to our understanding of the biological mechanism in the regeneration, hypothesis should firstly proposed in a way that can represent the real  mechanism as much as possible. Usually a real-life problem, e.g. the spinal cord regeneration, consists of complex dynamics and interactions with many other environmental factors to be considered; an elaborate description of the real cases using a overly complex mathematical model sometimes can be impossible; assumptions and hypothesis are often proposed to make necessary simplification and abstract for a final reasonably detailed model. Some qualitative models are simple but illustrative for the real-life observed patterns, e.g. Lotka-Volterra (L-V) equation for predator and prey models; they are usually analytically solvable and some properties are used to make predictions (e.g. steady states in ecosystem). If a more precise model is desired, usually more effects or interactions are considered and added as terms of equations to a re-constructed model, e.g. considering environmental factors, a changing environmental capacity and predation from other species in L-V model.

Models like L-V equation are mostly analytically solvable when analysing properties of the model, however, analytical solution for many other general models are not feasible, because of more complex dynamics, or the model itself is not a deterministic model. In the areas where solutions are not analytically feasible, computational techniques are usually applied for approximate solutions. Solvers for different models had been proposed to simulate the system or find roots for some spacial points, such as Explicit Runge-Kutta \cite{dormand1980family} and LSODA.

For the regeneration process that we wanted to model, differential equations are our preference. Non-linear ordinary differential equations are used in this dissertation, where the only independent variable is time $t$, denoting the post-lesion time (hours). Further models can also consider other models in other forms, e.g. stochastic models and partial differential equations. For ordinary differential equations (ODE) system, there are several available computational solvers using different algorithms, implemented in MATLAB\footnote[1] {\url{https://uk.mathworks.com/help/matlab/math/choose-an-ode-solver.html}}, Python\footnote[2]{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.solve_ivp.html}} and many other languages. These solvers help to simulate the trajectories of the dependent variables, given initial values and a set of fixed parameters. The goodness of `fit' then can be measured by compared the simulated data and observed real data.

Often further researches on the modelled problems require the model to be deterministic on its parameters. Parameters in dynamic system models usually have their practical significance, e.g. describing reaction speed, environmental capacity or the birth rate. There are many approaches to derive these parameters from data or experience. For a popular topic and a widely used models, the parameters' values can be read from previous literatures; however in many others cases, we need to find parameter's values specifically for our model, and numerical values for parameters are mostly preferred, as they suitable for practical analysis and comparison.

To find the best parameters' values given observed data, either optimisation-based approaches (e.g. least-square fit (LS), maximal likelihood estimation (MLE)) or inference-based approaches can be used. Optimisation-based approaches may face two major drawbacks: over-fitting and local optima \cite{ref:abcsysbio}. Bayesian inference approaches gain popularity in recent years, as it can relief the concerns in over-fitting and provide a measurement for the uncertainty in the inferred values. Local optima can still be a concern, and many efforts have be proposed for a more through explore of the parameter space to avoid local modes.

@@

Alternatively, as $p(D)$ is fixed once the observed data is given, the rule can be rewritten as 
\begin{align}
    \label{eq:bayes2}
    p(\theta|D) \propto l(D|\theta)p(\theta)
\end{align}
where $l$ represents the likelihood function. In other words, Eqn \ref{eq:bayes2} can be described as `posterior is proportional to the product of likelihood and prior', and here posterior of parameters is our target in parameter estimation task. Prior $p(\theta)$ should express our initial believes or estimates on the parameter values before observing the data and conducting experiments; likelihood function $l(D|\theta)$ is used to describe how well the parameter set $\theta$ explains the data $D$, in probability values. Once the data is given, $\theta$ is the only parameter in the likelihood function. Some optimisation-based methods try to find estimates of parameter set $\theta$ s.t. maximise the likelihood, however, such methods will struggle when facing a problem where it is hard to specify an analytical expression of likelihood function, let alone calculating and comparing values of it.

Based on Eqn \ref{eq:bayes2}, to find the posterior of parameter set, prior can be set according to the problem context and our understanding and cover a reasonable ranges; computational methods can be used to approximate likelihood function, e.g. Markov Chain Monte Carlo (MCMC) \cite{ref:MCMC}. As an alterative solution, Approximate Bayesian computation (ABC) approaches do not focus on the evaluation of likelihood function, but focused on simulating the data from the given model and observed data, then used the simulated data to approximate the desired posterior. ABC-rejection \cite{ABC_rejection} is a simple example of the ABC process, and further algorithms e.g. sequential Monte-Carlo (ABC SMC)\cite{Toni} were proposed to improve the efficiency.

@@

\begin{enumerate}
    \item Draw a sample $\theta^*$ from parameters' prior distribution. $\theta^*$ is a vector that consists values for each parameter
    \item Plug $\theta^*$ into the model, and use the model to generate simulated data $D^*$, where $D^*$ is of the same form as the observed data $D$, e.g. includes same summary statistics
    \item The discrepancy between the simulated data and observed data is measured by a distance function $d(D^*, D)$. For a pre-fixed threshold value $\epsilon$, if $d(D^*, D)<\epsilon$ then the sample $\theta^*$ is accepted
    \item Repeat 1, 2 and 3 until enough numbers of samples are accepted. The posterior distribution then can be approximated by these samples
\end{enumerate}

In our study of the regeneration model, full trajectories of the dependent variables are used as the data to be compared, as our goal is to find a model with a parameter set that can intuitively recover the real process as much as possible. The mean values at each time point for each dependent variable is calculated to form $D$, and for each sample, the ODE is accumulatively solved at the same time points and generate the simulated trajectories $D^*$. Using other statistics to summary the raw data are also possible but care must be taken as the choice of summary statistics can largely affect the resultant posterior \cite{summaryD, summaryD2}.

Instead of fixing the threshold value, the algorithm can be more adaptive by using a sequence of deceasing threshold values $\epsilon_t$, and derive new populations (called generation) of samples from the accepted samples (called particles) in previous population. An example is ABC SMC \cite{Toni}, which can significantly reduce the required samples compared to ABC-rejection \cite{ref:disease}:

@@

An illustrative graph of the ABC SMC procedures can be found in Figure \ref{fig:smc}. In implementations, many factors can affect the efficiency and goodness of fit, some of them are discussed in the Chapter 4. Rather than a pre-fixed threshold schedule, median or quantile based epsilon schedule can save the total sampling size and thus improve acceptance rates of particles \cite{threshold}. Some other improvements include using adaptive population size for each generation \cite{population}, using adaptive distance function \cite{ref:adpt_dis} and switching kernels \cite{ref:kernel}, etc. In our implementations these options were tried and compared when applicable and suitable.

@@

In addition, ABC SMC is also capable of model selection tasks \cite{model_compare}. Instead of draw samples from prior of a single model, we can draw samples from multiple models $m_i$ and approximate $p(m_i, \theta|D)$, which is the joint probability distribution of model $m_i$ and its accepted particles $\theta$. In each generation the marginal distribution $p(m_i|D)$ can be calculated as the accepted ${\theta}$ is known, thus models can be ranked according to their probabilities.

The feature that ABC algorithm can perform likelihood-free inference makes it widely used in many dynamic modeling problems, especially in the area of biology \cite{ref:abcsysbio, ref:disease, ref:compare}. But ABC is not limited to these areas, as it can be generally used in inference tasks to give parameter estimates or model rankings. E.g., it has been successfully applied in cosmology studies \cite{cosmology}.

@@

 A software packages that integrally implement ABC SMC is preferred, and more options in the settings of the algorithm, more customisable features make it better. Moreover, as the ABC SMC itself is a computationally intensive task that contains massive parallel potentials, an ideal software package is expected to support as least one kind of parallel options e.g. multi-core, GPU accelerating and clusters. Otherwise, a modification or rewrite of the software package are expected to produce a reliable input-output framework for our parameter estimation and model selection experiments, where much more work would be introduced.

 Preliminarily \verb|pyABC| packages \cite{ref:pyabc} in \verb|Python| was chosen for implementations. \verb|pyABC| supports multiple types of parallelisation, e.g., multicore processing and distributed cluster, and supports resume stored runs, which made our experiments easier. It is well documented and flexible in customisations of the algorithm.

 Some other ABC SMC related packages were also examined and considered as backups, include ELFI \footnote{\url{https://elfi.readthedocs.io/en/latest/}}, ABC-SysBio \cite{ref:abcsysbio}, EasyABC\footnote{\url{https://cran.r-project.org/web/packages/EasyABC/index.html}}, GpABC \footnote{\url{https://github.com/tanhevg/GpABC.jl}}, etc. These packages may have their advantages (e.g. ABC-SysBio supports CUDA acceleration) and considered as alternative implementation methods, or can be used to compare performance difference.

 Additionally, some other softwares were also need for solving ODE, results analysis and visualisations. These software and hardware environments are listed in Appendix A.

>>>>

Compared to mammals, zebrafish can regenerate the spinal cord after injury, and possibly recover swimming function. They are widely used in regenerative ability researches due to this ability. Experiments discovered that although being cut off, the axonal of nerve cells can regenerate and bridge across the lesion site, which critically determined whether zebrafish can regain the swimming function \cite{axonal}. Tsarouchas et al. \cite{ref:Tsarouchas} studied the complex interactions that happen at the lesion site. The inflammation affects the regeneration: if it is promoted, it accelerates the axonal regeneration; if it is inhibited, it reduces the regeneration\cite{ref:Tsarouchas}. Further experiments showed that two kinds of cytokines, il-1$\beta$ and tnf-$\alpha$, are related the inflammation, and they are dynamically controlled by immunes cells, i.e. immune cells produce and control the cytokines-mediated inflammation \cite{ref:Tsarouchas}. Cytokines are small proteins, work as messengers to pass signal across different cells, which are similar to hormones and neurotransmitter.

An interaction map (Figure \ref{fig:map}) can be drawn according to the evidences found in experiments.\cite{ref:Tsarouchas}. After the injury, a massive immune response was observed, with the increasing numbers of several kinds of cells. This dynamical system consists of complex interactions and dynamical feedback between cells and cytokines. For example, il-1$\beta$ promotes macrophages' increase. As a primary source, more macrophages leads to more expression of tnf-$\alpha$ mRNA.

Consequently, tnf-$\alpha$ and il-1$\beta$ plays determinative roles in the axonal regeneration. il-1$\beta$ promotes the regeneration at early stages, but later has an inhibition effect; tnf-$\alpha$ promotes the regeneration all the time. The presence of the cytokines also affects the number of immune cells, while as sources, more immune cells produce more cytokines. Interactions between different types of cells are also considered in the system (e.g., the number of neutrophils is affected by macrophages).

@@

    \caption[Working model of the influence of the innate immune system on axonal regrowth]%
    {Working model of the influence of the innate immune system on axonal regrowth, from Tsarouchas et al. \cite{ref:Tsarouchas}. Lines ended with arrows represent promoting effect, lines ended with T-connectors represent inhibition}
    \label{fig:map}

@@

This dynamic system is suitable for mathematical modelling, where a set of equations can be written to represent the dynamics of different objects, e.g. each type of cells and cytokines. Our interest lies in the modelling of the regeneration period of the dynamic system, and we wish to find that if we can quantitatively study the interactions inside the system, based on the proposed hypothesis \cite{ref:Tsarouchas},  furthermore based our hypothesis which may include different interactions.

The biological background of works in this dissertation is mostly related to the experimental evidence, quantitative data and proposed hypothesis in \cite{ref:Tsarouchas}. The published data on the quantitative measurement of cells and cytokines make the modelling and further analysis possible. Some other interactions proposed in our modelling stages are enlightened from Anderson et al. \cite{Anderson}, where intracellular cytokine signalling networks were established and computationally modelled.

@@

Mathematical models can express real-life problems in mathematical forms, and provide qualitative or quantitative inside-look of the question through various mathematical tools and techniques. They can be used for prediction, classification, testing our understanding or hypothesis and many other tasks; thus, they are widely used to solve practical problems. However, for a model that describes the zebrafish spinal cord regeneration process, we were less interested in the model's predictive functionality; the main focus is on the exploration of uncertainties in parameter estimates and comparison with alternative models. In this case, we wish to use modelling techniques and implement inference methods, test our hypothesis and hopefully find and compare well-fitted models.

According to our understanding of the biological mechanism in the regeneration, hypotheses should be proposed in a way that can represent real mechanism as much as possible. Usually, a real-life problem, e.g. the spinal cord regeneration, consists of complex dynamics and interactions with many other environmental factors to be considered. An elaborate description of the real cases using an overly complicated mathematical model sometimes can be impossible; assumptions and hypotheses are often proposed to make necessary abstract and simplification for a final reasonably detailed model. Some qualitative models are illustrative but straightforward for the observed patterns, e.g. Lotka-Volterra (L-V) equation for predator and prey models. They are usually analytically solvable, and their properties are used to make predictions (e.g. steady states in the ecosystem). If a more precise model is desired, usually more influences or interactions are considered and added as terms of equations to a re-constructed model, e.g. considering environmental factors, a changing environmental capacity and predation from other species in L-V model.

Models like the L-V equation are mostly analytically solvable when analysing properties of the model. However, analytical solutions for many other general models are not feasible because of more complicated dynamics, or the model itself is not a deterministic model. In the areas where solutions are not analytically feasible, computational techniques are usually applied for approximate solutions. Solvers for different models had been proposed to simulate the system or find roots for some special points, such as Explicit Runge-Kutta \cite{dormand1980family} and LSODA.

For the regeneration process that we wanted to model, differential equations are our preference. Non-linear ordinary differential equations are used in this dissertation, where the only independent variable is time $t$, denoting the post-lesion time (hours). Further models can also consider other models in other forms, e.g. stochastic models and partial differential equations. For ordinary differential equations (ODE) system, there are several available computational solvers using different algorithms, implemented in MATLAB\footnote[1] {\url{https://uk.mathworks.com/help/matlab/math/choose-an-ode-solver.html}}, Python\footnote[2]{\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.integrate.solve_ivp.html}} and many other languages. These solvers help to simulate the trajectories of the dependent variables, given initial values and a set of fixed parameters. The decency of fit can be measured by compared the simulated data and observed real data.

Often further researches on the modelled problems require the model to be deterministic on its parameters. Parameters in dynamic system models usually have their practical significance, e.g. describing reaction speed, environmental capacity or the birth rate. There are many approaches to derive these parameters from data or experience. For a popular topic and widely used models, parameters' values can be read from previous literature. However, in many other cases, we need to find values explicitly for our model, where numerical values for parameters are mostly preferred as they are suitable for practical analysis and comparison.

To find the best parameters' values given observed data, either optimisation-based approaches (e.g. least-squares fit (LS), maximal likelihood estimation (MLE)) or inference-based approaches can be applied. Optimisation-based approaches may face two major drawbacks: over-fitting and local optima \cite{ref:abcsysbio}. Bayesian inference approaches gain popularity in recent years, as it can relieve the concerns in over-fitting and provide a measurement for the uncertainty in the inferred values. Local optima can still be a concern, and many efforts have been proposed for a more thorough exploration of the parameter space to avoid local modes.

@@

Alternatively, as $p(D)$ is fixed once the observed data is provided, the rule can be rewritten as 
\begin{align}
   \label{eq:bayes2}
   p(\theta|D) \propto l(D|\theta)p(\theta)
\end{align}
where $l$ represents the likelihood function. In other words, Eqn. \ref{eq:bayes2} can be described as `posterior is proportional to the product of likelihood and prior', and here posterior of parameters is our target in the parameter estimation task. Prior $p(\theta)$ should express our initial believes or estimates on the parameter values before observing the data and conducting experiments; likelihood function $l(D|\theta)$ is used to describe how well the parameter set $\theta$ explains the data $D$, in probability values. Once the data is given, $\theta$ is the only parameter in the likelihood function. Some optimisation-based methods attempt to find estimates of parameter values $\theta$ s.t. maximise the likelihood. However, such methods will struggle when facing a problem where it is hard to specify an analytical expression for likelihood function, let alone calculating and comparing values of it.

Based on Eqn. \ref{eq:bayes2}, to find the posterior of the parameter set, prior can be set according to the problem context and our understanding and cover a reasonable range. Computational methods can be used to approximate likelihood function, e.g. Markov Chain Monte Carlo (MCMC) \cite{ref:MCMC}. As an alternative solution, Approximate Bayesian computation (ABC) approaches do not focus on the evaluation of likelihood function but focus on simulating the data from the given model and observed data, then use the simulated data to approximate the true posterior. ABC-rejection \cite{ABC_rejection} is a simple example of the ABC process, and further algorithms e.g. sequential Monte-Carlo (ABC SMC)\cite{Toni} were proposed to improve the efficiency.

@@

\begin{enumerate}
    \item Draw a sample $\theta^*$ from parameters' prior distribution. $\theta^*$ is a vector that consists of values for each parameter
    \item Plug $\theta^*$ into the model, and use the model to generate simulated data $D^*$, where $D^*$ is of the same form as the observed data $D$, e.g. includes same summary statistics
    \item The discrepancy between the simulated data and observed data is measured by a distance function $d(D^*, D)$. For a pre-fixed threshold value $\epsilon$, if $d(D^*, D)<\epsilon$ then the sample $\theta^*$ is accepted
    \item Repeat 1, 2 and 3 until enough numbers of samples are accepted. The posterior distribution then can be approximated by these samples
\end{enumerate}

In our study of the regeneration model, full trajectories of the dependent variables are used as the data to be compared, as our goal is to find a model with a parameter set that can intuitively recover the real process as much as possible. The mean values at each time point for each dependent variable are calculated to form $D$. For each sample, the ODE is accumulatively solved at the same time points and generate the simulated trajectories $D^*$. Using other statistics to summarise the raw data are also possible, but care must be taken as the choice of summary statistics can largely affect the resultant posterior \cite{summaryD, summaryD2}.

Instead of fixing the threshold value, the algorithm can be more adaptive by using a sequence of decreasing threshold values $\epsilon_t$, and derive new populations (called generation) of samples from the accepted samples (called particles) in the previous population. An example is ABC SMC \cite{Toni}, which can significantly reduce the required samples compared to ABC-rejection \cite{ref:disease}:

@@

An illustrative graph of the ABC SMC procedures can be found in Figure \ref{fig:smc}. In implementations, many factors can affect the efficiency and goodness of fit; some of them are discussed in Chapter 4. Rather than a pre-fixed threshold schedule, median or quantile based epsilon schedule can save the total sampling size and thus improve acceptance rates of particles \cite{threshold}. Some other improvements include using adaptive population size for each generation \cite{population}, using adaptive distance function \cite{ref:adpt_dis} and switching kernels \cite{ref:kernel}, etc. In our implementations, these options were tried and compared when applicable.

@@

In addition, ABC SMC is also capable of model selection tasks \cite{model_compare}. Instead of draw samples from prior of a single model, we can draw samples from multiple models $m_i$ and approximate $p(m_i, \theta|D)$, which is the joint probability distribution of model $m_i$ and its accepted particles $\theta$. In each generation, the marginal distribution $p(m_i|D)$ is calculated as the accepted ${\theta}$ is known; thus models can be ranked according to their probabilities.

The feature that ABC algorithm can perform likelihood-free inference makes it widely used in many dynamic modelling problems, especially in the area of biology \cite{ref:abcsysbio, ref:disease, ref:compare}. However, ABC is not restricted to these areas, as it can be generally used in inference tasks to give parameter estimates or model rankings. E.g., it has been successfully applied in cosmology studies \cite{cosmology}.

@@

Software packages that integrally implement ABC SMC are preferred, and more options in the settings of the algorithm, more customisable features make it better. Moreover, as the ABC SMC itself is a computationally intensive task that contains massive parallel potentials, an ideal software package is expected to support as least one kind of parallel options, e.g. multi-core, GPU accelerating and clusters. Otherwise, a modification or rewrite of the software package is expected to produce a reliable input-output framework for our parameter estimation and model selection experiments, where much more work would be introduced.

Preliminarily \verb|pyABC| packages \cite{ref:pyabc} in \verb|Python| was chosen for implementations. \verb|pyABC| supports multiple types of parallelisation, e.g., multi-core processing and distributed cluster, and supports resume stored runs, which made our experiments easier. It is well documented and flexible in customisations of the algorithm.

Some other ABC SMC related packages were also examined and considered as backups, include ELFI \footnote{\url{https://elfi.readthedocs.io/en/latest/}}, ABC-SysBio \cite{ref:abcsysbio}, EasyABC\footnote{\url{https://cran.r-project.org/web/packages/EasyABC/index.html}}, GpABC \footnote{\url{https://github.com/tanhevg/GpABC.jl}}, etc. These packages may have their advantages (e.g. ABC-SysBio supports CUDA acceleration) and considered as alternative implementation methods, or can be used to compare performance difference in future works.

Additionally, some other software tools were also needed in ODE solving, result analysis and visualisations. The software and hardware environments used in this dissertation are listed in Appendix A.

### 2020-08-16 introduction/introduction.tex
Grammar fix

Zebrafish has the ability to regenerate its spinal cord after injury; tissue regeneration observed at the injury site is a complex biological process with interacting parts and dynamical feedback between different types of cells and molecules. These cells and molecules are involved in the inflammation and repair in a dynamical way and a complex interacting network can be drawn according to it.

@@

Mathematical models and techniques can help us to resolve the complexity in this dynamic system, however, how to derive the parameter estimates of the models remains a challenging work. In this dissertation, our interest lies in the modeling of the regeneration process that happens around the lesion site, especially in the parameter estimates and comparison of possible models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was firstly proposed but the addressed interactions remains to be precisely defined, and quantitatively examined in experiments. We would like to find the best parameter sets of the model using the published data, and ranking models under certain metrics and possibly find the best-fit model. In these steps, models are expected to retain mathematical and biological significance regarding the complex interactions and differential equations are suitable for this.

@@

Fo the parameter estimation tasks, optimisation-based methods usually have trouble in determining global optima from local optimal, and noise in the observed data may lead to over-fitting. Compared to optimisation-based methods, Bayesian inference techniques can help to relief these two concerns\cite{ref:abcsysbio}. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression for the dynamic system can be hard.

Such methods introduce computation-intensive sampling-and-test patterns, where massive number of samples are drawn and followed by complex computations. As the sampling is independent from each other in a certain period, the parallel implementation of the algorithm becomes feasible and is expected to accelerate the parameter inference in a large scale when using HPC resources.

This dissertation mainly focuses on the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. Additionally, performance experiments illustrated the strong-scaling performance of the algorithm. Chapter 2 talks about the background information and theories, Chapter 3 shows how the mathematical models are derived from hypothesis and Chapter 4 describes the ABC SMC implementations and experiments on the models in detail. A brief scaling-up performance experiment is included in Chapter 5, and Chapter 6 summarises the conclusions and states the future works on this topic.

>>>>

Zebrafish can regenerate its spinal cord after injury; tissue regeneration observed at the injury site is a complex biological process with interacting parts and dynamical feedback between different types of cells and molecules. These cells and molecules are involved in the inflammation and repair dynamically, and a complex interacting network can be drawn according to it.

@@

Mathematical models and techniques can help us to resolve the complexity in this dynamic system; however, how to derive the parameter estimates of the models remains a challenging work. In this dissertation, our interest lies in the modelling of the regeneration process that happens around the lesion site, especially in the parameter estimates and comparison of possible models. According to available researches \cite{ref:Tsarouchas}, a mathematical model was first proposed, but the addressed interactions remain to be precisely defined, and quantitatively examined in experiments. We want to find the best parameter sets of the model using the published data, and ranking models under specific metrics and possibly find the best-fit model. In these steps, models are expected to retain mathematical and biological significance regarding the complex interactions, and differential equations are suitable for this.

@@

For the parameter estimation tasks, optimisation-based methods usually have trouble in determining global optima from local optimal, and noise in the observed data may lead to over-fitting. Compared to optimisation-based methods, Bayesian inference techniques can help to relieve these two concerns\cite{ref:abcsysbio}. Moreover, approximate inference methods (Approximate Bayesian Computation, ABC) make likelihood-free inference possible, which is suitable for our task as writing down likelihood as a function expression for the dynamic system can be hard.

Such methods introduce computation-intensive sampling-and-test patterns, where a massive number of samples are drawn and followed by complex computations. As the sampling is independent of each other in a certain period, the parallel implementation of the algorithm becomes feasible and is expected to accelerate the parameter inference on a large scale when using HPC resources.

This dissertation mainly focuses on the implementations and experiments of parameter estimation using ABC SMC (sequential Monte-Carlo) method, and discussions on the results. Additionally, performance experiments illustrated the strong-scaling performance of the algorithm. Chapter 2 talks about the background information and theories, Chapter 3 shows how the mathematical models are derived from hypothesis, and Chapter 4 describes the ABC SMC implementations and experiments on the models in detail. A brief scaling-up performance experiment is included in Chapter 5, and Chapter 6 summarises the conclusions and states the future works on this topic.

### 2020-08-16 performance/performance.tex
Grammar fix

    \caption[TODO]{Speedup. MORE DESCRIPTION}
    \label{fig:speedup}
\end{figure}

@@

    \caption[TODO]{Samples per second. MORE DESCRIPTION}
    \label{fig:sample_per_sec}
\end{figure}

@@

    \caption[TODO]{Local modes. MORE DESCRIPTION}
    \label{fig:local_modes}
\end{figure}

@@

    \caption[TODO]{Parameter density distribution. MORE DESCRIPTION}
    \label{fig:local_para}
\end{figure}

>>>>

    \caption[Speedup of the program]{Speedup of the program}
    \label{fig:speedup}
\end{figure}

@@

    \caption[Average samples in a second, under different number of cores]{Average samples in one second, under different number of cores}
    \label{fig:sample_per_sec}
\end{figure}

@@

    \caption[Local optimum and abnormal pattern]{(a) Required sampling size for two example runs. Different colour represents different generation (from bottom to top: generation 1 to generation 20). Run A took huge amount of time in population 11. (b) Required sample in each generation of Run A. (c) Time-consumption in each generation of Run A}
    \label{fig:local_modes}
\end{figure}

@@

    \caption[Parameter density distribution before and after the concentration movement in generation 11]{Parameter density distribution before (generation 9, cyan) and after (generation 13, red) the concentration movement in generation 11}
    \label{fig:local_para}
\end{figure}
